Errorless irrationality: removing error-driven components from the inverse base-rate effect paradigm

### 2023-01-13 docs/manuscript.tex

The inverse base-rate effect is a robust irrational bias that arises when people face ambiguity.
The most prominent theories of this irrational bias depend on prediction error.
In this study, we gradually removed elements of a predictive learning design to test the extent to which error-driven processes underly this bias.
In our first experiment, we removed explicit feedback by implementing the inverse base-rate effect in an observational learning procedure. 
In our second study, we further removed any causal relationship between stimulus features and category labels by moving towards an unsupervised learning procedure.
This removed any information participants could use to identify category labels.
In both experiments, the inverse base-rate effect persisted and remained robust.
This outcome suggests that this irrational bias is independent of supervised learning procedures.
We propose that any theories and models of the inverse base-rate effect must manage information encoding and connection updates without explicit prediction error.
We end by proposing two clear paths for future investigations.

@@

In the test phase, participants judged individual symptoms and novel combinations of old symptoms, see Table \ref*{tab:abstract-exp1}.
Symptoms appeared in a sentence, such as '\textit{John has a fever.}', with a prompt asking participants to label what disease the person has, '\textit{Does the patient has disease Z or disease L?}'.
Participants had to respond by pressing either Z or L on the keyboard.
They had 10 seconds to do so, otherwise, a '\textit{Please respond faster!}' message appeared.
After the button press, there was no feedback,
Each unique test item and training item (occurring in the test phase) was repeated 20 times.
So, the test phase included 120 trials, which were broken down into 5 blocks of 24 trials.

@@

In this experiment, we implemented the IBRE in a way most similar to cued-recall tasks.
Previous category labels were treated as features.
And features were selected to be solid black geometric shapes.
The task asked participants to memorise the arrangement of these shapes.
On each trial, we randomised the position of the geometric shapes in the arrangement.
This further minimised the chances of having any design component suggestive of which feature is the category label.

@@

% Asymmetric representation can arise in different ways.
One hypothesised way asymmetric representation is manifested is the attentional tuning of cognitive representation of category exemplars.
This is not necessarily absent in our experiments but is not directly tested.
Our experiments do not give direct evidence against the role of attention in developing asymmetric representation.
Nonetheless, it must not happen through an error-driven process as conceptualised in most prominent theories.
To further investigate this, the cued-recall procedure could incorporate eye-tracking to measure dwell time and order of information encoding.
Even if we observe more and longer fixations to C relative to B, attention can still be allocated asymmetrically to drive the distinguishability of categories.
Explanations do not need to invoke an explicit error-driven process.

>>>>

The inverse base-rate effect is a robust irrational bias that arises when people face ambiguity.
The most prominent theories of this irrational bias depend on prediction error.
In this study, we gradually detached elements of a predictive learning design to test the extent to which error-driven processes underly this bias.
In our first experiment, we removed explicit feedback by implementing the inverse base-rate effect in an observational learning procedure. 
In our second study, we further removed any causal relationship between stimulus features and category labels by moving towards an unsupervised learning procedure.
This removed any information participants could use to identify category labels.
In both experiments, the inverse base-rate effect persisted and remained robust.
This outcome suggests that this irrational bias is independent of supervised learning procedures.
We propose that any theories and models of the inverse base-rate effect must manage information encoding and connection updates without explicit prediction error.
We end by proposing two clear paths for future investigations.

@@

In the test phase, participants judged individual symptoms and novel combinations of old symptoms, see Table \ref*{tab:abstract-exp1}.
Symptoms appeared in a sentence, such as '\textit{John has a fever.}', with a prompt asking participants to label what disease the person has, '\textit{Does the patient has disease Z or disease L?}'.
Participants had to respond by pressing either Z or L on the keyboard.
They had 10 seconds to do so, otherwise, a '\textit{Please respond faster!}' message appeared.
After the button press, there was no feedback,
Each unique test item and training item (occurring in the test phase) was repeated 20 times.
The test phase, therefore, included 120 trials, which were broken down into 5 blocks of 24 trials.

@@

In this experiment, we implemented the IBRE in a way most similar to unsupervised category learning paradigms.
Previous category labels were treated as features.
And features were selected to be solid black geometric shapes.
The task asked participants to memorise the arrangement of these shapes.
On each trial, we randomised the position of the geometric shapes in the arrangement.
This further minimised the chances of having any design component suggestive of which feature is the category label.

@@

% Asymmetric representation can arise in different ways.
One hypothesised way asymmetric representation is manifested is the attentional tuning of cognitive representation of category exemplars.
This is not necessarily absent in our experiments but is not directly tested.
Our experiments do not give direct evidence against the role of attention in developing asymmetric representation.
Nonetheless, it must not happen through an error-driven process as conceptualised in most prominent theories.
To further investigate this, the unsupervised IBRE procedure could incorporate eye-tracking to measure dwell time and order of information encoding.
Even if we observe more and longer fixations to C relative to B, attention can still be allocated asymmetrically to drive the distinguishability of categories.
Explanations do not need to invoke an explicit error-driven process.
